Predicting the effects of character displacement on food-web dynamics

### 2019-10-30 Appendix_S4.tex
finalize edits for Appendix S4 for bioRxiv submission

%%% From AmNat_MS_template.tex
%\documentclass[11pt]{article}
%\usepackage[sc]{mathpazo} %Like Palatino with extensive math support
\usepackage{fullpage}
%\usepackage[authoryear,sectionbib,sort]{natbib}
\linespread{1.5} % 1.7
%\usepackage[utf8]{inputenc}
\usepackage{lineno}

% controlling text size in figure captions
\usepackage{caption}
\captionsetup[figure]{font=small}

%%%%%%% Sections not currently working properly
%\usepackage{titlesec}
%\titleformat{\section}[block]{\Large\bfseries\filcenter}{\thesection}{1em}{}
%\titleformat{\subsection}[block]{\Large\itshape\filcenter}{\thesubsection}{1em}{}
%\titleformat{\subsubsection}[block]{\large\itshape}{\thesubsubsection}{1em}{}
%\titleformat{\paragraph}[runin]{\itshape}{\theparagraph}{1em}{}[. ]\renewcommand{\refname}{Literature Cited}

%%% My added packages
%\usepackage{amsmath} % for matrices
%\usepackage{graphicx} % for organizing figures
%\usepackage{lmodern} % for tilde
%\usepackage[T1]{fontenc} % for tilde

%%%%%%%%% Previous - likely useful though
%\usepackage{setspace}\doublespacing
%\usepackage{float}
%\let\origfigure\figure
%\let\endorigfigure\endfigure
%\renewenvironment{figure}[1][2] {
%    \expandafter\origfigure\expandafter[H]
%} {
%    \endorigfigure
%}

@@

\begin{center} \LARGE Ecological character displacement destabilizes food webs \end{center}

@@

\emph{Running title}: Character divergence destabilizes food webs

\emph{Keywords}: competition; eco-evolutionary dynamics;
consumer-resource interactions; adaptation; community stability

\linenumbers{} \modulolinenumbers[3]

@@

\section{Abstract}\label{abstract}

Ecological character displacement is an adaptive process that generally
increases phenotypic diversity. Despite the fact that this
diversification is due to an eco-evolutionary feedback between consumers
competing for shared resources, its consequences for food-web dynamics
have not been examined. Here, I study a model of two consumers competing
for two shared resources to examine how character displacement in
consumer attack rates affects resource abundances and the resilience of
food webs to perturbations. I found that character displacement always
strengthened consumer-resource interactions whenever consumers competed
for resources that occurred in different habitats. This increase in
interaction strength resulted in lower resource abundances and less
resilient food webs. This occurred under different evolutionary
tradeoffs and in both simple and more realistic foraging scenarios.
Taken together, my results show that the adaptive process of character
displacement may come with the ecological cost of decreasing food-web
resilience.

@@

\section{Introduction}\label{introduction}

Ecological character displacement is an important adaptive process in
generating biodiversity \citep{Schluter2000, Pfennig2010}. This process
is due to ``phenotypic evolution in a species generated or maintained by
{[}exploitative{]} resource competition with one or more coexisting
species'' \citep{Schluter2000}. A large body of theoretical
\citep[e.g.][]{Lawlor1976, Abrams1986, Doebeli1996, Taper1985, McPeek2019}
and empirical \citep[reviewed
in:][]{Schluter2000, Dayan2005, Stuart2013} work has examined which
scenarios lead to phenotypic divergence or convergence of competing
consumers. The general conclusion has been that, if resources are
nutritionally substitutable \citep{Abrams1987, Fox2008} and there is no
other strong source of density dependence acting on consumers
\citep{Abrams1986}, then resource competition drives the adaptive
divergence of competitors \citep{Lawlor1976, Taper1985}. This adaptive
process is not simply a response to static differences in resource
distributions, but creates an eco-evolutionary feedback that drives
further differentiation. This crucial insight was made by theoretical
models that explicitly included resource dynamics as a mediator of
competition in driving evolutionary change
\citep{Lawlor1976, Abrams1986, Taper1985}.

Although models that included resources led to insights to the evolution
of character displacement, the ecological feedback onto
consumer-resource dynamics has received surprisingly little attention.
This is likely because the ecological feedback has been primarily
studied through the lens of coexistence theory
\citep{Lawlor1976, Germain2018, Bassar2017, McPeek2019}. For example,
early theoretical work showed that ecological character displacement
promotes coexistence by favoring specialized consumers that experience
reduced interspecific competition \citep{Lawlor1976}. Yet, this
reduction in interspecific competition may, at the same time, increase
interspecific interactions between specialized consumers and their
resources. Both food-web theory and empirical studies have shown that
increasing the strength of consumer-resource interactions often
suppresses the abundance of resources, which if sufficient enough, can
generate oscillations and less stable consumer-resource dynamics
\citep{Rosenzweig1971, Luckinbill1973, Murdoch2002, Murdoch2003, McCann2011}.
Thus, a food-web perspective, which accounts for both the direct and
indirect effects of consumer-resource interactions, may yield new
insight to the ecological consequences of character displacement.

Here, I address this knowledge gap by studying a mathematical model that
examines how ecological character displacement affects consumer-resource
dynamics in a food-web context. I address two questions: (1) How does
ecological character displacement affect resource abundances? (2) How
does character displacement affect food-web stability? To test the
generality of these effects, I explored different ecological foraging
scenarios and evolutionary tradeoffs in consumer attack rates. I found
that the adaptive process of character displacement often comes with an
ecological cost; resulting in food webs with lower resource availability
and that are less resilient to perturbations.

\section{Material and methods}\label{material-and-methods}

\subsection{Underlying Consumer-Resource
Dynamics}\label{underlying-consumer-resource-dynamics}

To examine how ecological character displacement affects resource
abundances and food-web stability, I analyzed a continuous-time model of
two consumers (\(C_{j=1,2}\)) competing for two shared resources
(\(R_{i=1,2}\)):

\begin{equation} \label{eq:1}
  \begin{split}
     & \frac{dR_1}{dt}=r_1R_1(1-\frac{R_1}{K_1})-F_{11}(R_1)C_1-F_{12}(R_1)C_2 \\
     & \frac{dR_2}{dt}=r_2R_2(1-\frac{R_2}{K_2})-F_{21}(R_2)C_1-F_{22}(R_2)C_2 \\
     & \frac{dC_1}{dt}=e_{11}F_{11}(R_1)C_1+e_{21}F_{21}(R_2)C_1-m_1C_1 \\
     & \frac{dC_2}{dt}=e_{12}F_{12}(R_1)C_2+e_{22}F_{22}(R_2)C_2-m_2C_2 \\
  \end{split}
\end{equation}

where \(r_i\) represents the intrinsic growth rate of resource \(i\),
\(K_i\) represents the carrying capacity of resource \(i\), \(e_{ij}\)
represents the conversion efficiency of resource \(i\) into consumer
\(j\), and \(m_j\) represents the mortality rate of consumer \(j\).
\(F_{ij}(R_i)\) represents consumer \(j\)'s feeding rate on resource
\(i\) (i.e.~functional response). This model is a useful
characterization of a scenario where consumers compete for two distinct
resources (e.g.~zooplankton and benthic invertebrates in lakes) rather
than a scenario where resources are better characterized by a continuous
trait distribution (e.g.~seed size, see \citet{Taper1985} for an
example). Importantly, inferences about ecological character
displacement can only be made by comparing food webs with and without a
competing consumer \citep{Schluter1992}. Therefore, I arbitrarily set
\(C_2=0\) to create a food-web without a competing consumer for these
comparisons.

\subsection{Foraging Scenarios}\label{foraging-scenarios}

I studied three different foraging scenarios. In the first, I assume
that consumers can forage for both resources simultaneously (fig.
\ref{fig:foraging_scenarios} top) and their feeding rates increase
linearly with resource abundance, such that:

\begin{equation} \label{eq:2}
  F_{ij}(R_i)=a_{ij}R_i
\end{equation}

where \(a_{ij}\) is the attack rate of consumer \(j\) on resource \(i\).
This first scenario is the starting point for many models of resource
competition \citep{MacArthur1972}; however, it does not reflect many
food webs where consumers are mobile and their foraging behavior links
resources that occur in different habitats \citep{McCann2005}. The
second scenario accounts for this spatial context (fig.
\ref{fig:foraging_scenarios} bottom) and takes the form:

\begin{equation} \label{eq:3}
  F_{ij}(R_i)=w_{ij}a_{ij}R_i
\end{equation}

where \(w_{ij}\) represents the proportion of time consumer \(j\) spends
foraging in a habitat where only resource \(i\) is found (i.e.~habitat
preference). Note that since \(w_{ij}\) is a proportion that
\(w_{1,j}=1-w_{2,j}\). Finally, it is well known that consumer feeding
rates often saturate at high resource abundances
\citep{Holling1959, Rosenzweig1963, Murdoch2003, McCann2011} and that
consumers do not usually spend a fixed proportion of time in a
particular habitat \citep{McCann2005}. The third scenario accounts for
these biological realities and takes the form \citep[derived
by][]{McCann2005}:\\

\begin{equation} \label{eq:4}
  F_{ij}(R_i)=\frac{a_{ij}W_{ij}R_i}{1+a_{1,j}h_{1,j}W_{1,j}R_1+a_{2,j}h_{2,j}W_{2,j}R_2}
\end{equation}

where consumer \(j\)'s feeding rate on resource \(i\) is influenced by
the abundance of each resource; saturates as resource abundances
increase (due to handling times \(h_{ij}\)); and consumer habitat
preferences are modified by the relative abundance of resources, such
that: \(W_{ij}=\frac{w_{ij}R_i}{w_{1,j}R_1+w_{2,j}R_2}\).

Previous studies have analyzed the evolution of consumer attack rates in
the first two foraging scenarios using an Adaptive Dynamics approach,
with the general result being character divergence
\citep{Lawlor1976, Abrams1986}. I also used an Adaptive Dynamics
approach to analyze character displacement in the third foraging
scenario, and I too observed character divergence (detailed analysis in
Appendix S1 of Supplementary Information). I say consumers have
undergone character divergence if their evolved attack rates are more
specialized when evolving with vs.~without a competing consumer.
Specialization of consumer \(j\) on resource \(1\) is measured as
\(\frac{a_{1,j}}{a_{1,j}+a_{2,j}}\), where a value of 0.5 is a complete
generalist (\(a_{1,j}=a_{2,j}\)), and a value of 1 is a complete
specialist (\(a_{2,j}=0\)). Values less than 0.5 indicate specialization
on the other resource.

\begin{figure}
\centering
\includegraphics{ECD_model_foraging_scenarios}
\caption{\label{fig:foraging_scenarios}\textbf{Ecological foraging
scenarios.} I examined whether the effect of ecological character
displacement on food-web dynamics depended on whether consumers competed
for resources that occurred in the same vs.~different habitats. The
width of each arrow corresponds to initial attack rates
(\emph{a}\textsubscript{ij}) of consumer \emph{j} on resource \emph{i}.
Note that \emph{C}\textsubscript{1} was pre-adapted to
\emph{R}\textsubscript{1} (\emph{a}\textsubscript{11} \textgreater{}
\emph{a}\textsubscript{21}), while \emph{C}\textsubscript{2} was a
mirror image, being pre-adapted to \emph{R\textsubscript{2}}
(\emph{a}\textsubscript{22} \textgreater{} \emph{a}\textsubscript{12}).
In each scenario, I assumed consumer feeding rates increased linearly
with resource abundance. I also relax this assumption and consider a
more realistic functional response when resources occurred in different
habitats.}
\end{figure}

\subsection{Food-web Dynamics}\label{food-web-dynamics}

Given that character divergence occurred across these foraging
scenarios, I focus here on its consequences for food-web dynamics. To do
this, I analyzed differences in resource abundances and food-web
stability at equilibrium. An equilibrium is reached when the rates of
change in equation \ref{eq:1} are 0, and solving the system at this
point gives equilibrium abundances for each resource (\(\hat R_i\)) and
each consumer (\(\hat C_j\)). I also compared the local stability of
these food webs using standard methods \citep{Otto2007}. This stability
analysis derives the dominant eigenvalue, \(\lambda_{\text{max}}\), of
the matrix of partial derivatives of all differential equations with
respect to all variables evaluated at equilibrium. If
\(-\lambda_{max}>0\), then the food web will return to equilibrium after
a small perturbation (i.e.~locally stable), with more positive values
indicating a faster return time. If \(-\lambda_{max}<0\), then the food
web is not locally stable.

When possible, I derived analytical expressions for the relationship
between consumer attack rates and food-web dynamics. To do this, I
simplified the model by assuming that resources are equivalent
(\(r=r_i\) and \(K=K_i\)) as well as consumers (\(e=e_{ij}\);
\(h=h_{ij}\); \(m=m_j\)), except that consumer attack rates and their
habitat preferences (if present) are mirror images of each other
(\(a_{11}=a_{22}\); \(a_{12}=a_{21}\); \(w_{11}=w_{22}\)). Note that I
arbitrarily set \(C_1\) as being pre-adapted to \(R_1\)
(\(a_{11} > a_{21}\); \(w_{11} > 0.5\)), and therefore \(C_2\) was
pre-adapted to \(R_2\). Controlling for other sources of variability
allowed me to isolate the general effects of character divergence. All
mathematical derivations were conducted in Mathematica
\citep{Mathematica} and are provided in the Supplementary Information
(Appendix S1-3).

To gain insight to the eco-evolutionary feedback generated by character
displacement, I conducted simulations using an Adaptive Dynamics
approach. Specifically, after letting consumer and resource abundances
reach a steady state, I created a mutant consumer by randomly choosing
one and modifying its attack rate on one resource by either subtracting
or adding a small constant (0.01 in the following simulations) with
equal probability. The mutant's attack rate on the other resource was
determined by a tradeoff, such that
\(\big(a_{\text{1},j}/{A}\big)^n+\big(a_{\text{2},j}/{A}\big)^n=1\),
where \(A\) is the total investment in attack rates and \(n\) describes
the shape of the tradeoff \citep{Sargent2006}. This function has the
useful property that it differentiates between cases where intermediate
combinations of \(a_{\text{1},j}\) and \(a_{\text{2},j}\) are higher
than the extremes (when \(n>1\), green line in fig. \ref{fig:tradeoff})
or, conversely, where the two extremes are higher than intermediate
investments (when \(n<1\), orange line fig. \ref{fig:tradeoff}). When
\(n=1\), the tradeoff function is linear, and all combinations of
\(a_{\text{1},j}\) and \(a_{\text{2},j}\) have the same total attack
rate (blue line in fig. \ref{fig:tradeoff}). Assuming the mutant
consumer was rare, I then determined whether the mutant had higher
relative fitness than the resident consumer, and thus could invade and
replace the resident consumer. If the mutant was able to invade, I
updated the attack rate of the resident consumer to the mutant attack
rate and allowed consumer and resource abundances to reach a steady
state. I then repeated the simulation up to 10,000 times, which was
sufficient for consumers to either reach an evolutionary stable strategy
\citep[ESS,][]{Smith1973} or an evolutionary limit (e.g.
\(\frac{a_{ij}}{a_{\text{1},j}+a_{\text{2},j}}\) is constrained to a
maximum of 1 and minimum of 0). Unless otherwise noted, I conducted
simulations with the following parameter values: \emph{r} = 1; \emph{K}
= 4; \emph{e} = 0.8; \emph{m} = 1; \emph{A} = 2; \emph{h} = 0.4; and
\(w_{11} = w_{22} = 0.6\). I set an initial value of
\(a_{11} = a_{22} = 1.2\), while \(a_{12}\) and \(a_{21}\) depended on
the value of \emph{n}. I set initial consumer and resource abundances
to: \(R_1 = R_2 = 2\); \(C_1 = C_2 = 1\). All simulations were conducted
in R \citep{R} and the code to reproduce these simulations is publically
available on GitHub (\url{https://github.com/mabarbour/ECD_model}).

\begin{figure}
\centering
\includegraphics{attack_tradeoffs.pdf}
\caption{\label{fig:tradeoff}\textbf{Evolutionary tradeoffs in consumer
attack rates.} In each foraging scenario, I explored the effects of
three different tradeoffs: intermediate combinations of attack rates
(\emph{a}\textsubscript{1,\emph{j}}, \emph{a}\textsubscript{2,\emph{j}})
are higher than the extremes (green line, \emph{n} \textgreater{} 1);
extreme combinations of attack rates are higher than intermediate
investments (orange line, \emph{n} \textless{} 1); and all combinations
of attack rates have the same total attack rate (blue line, \emph{n} =
1). Points corresponding to attack rates at the beginning of the
simulation for \emph{C}\textsubscript{1}, which was pre-adapted to
\emph{R}\textsubscript{1} (\emph{a}\textsubscript{11} \textgreater{}
\emph{a}\textsubscript{12}).}
\end{figure}

\section{Results}\label{results}

\subsection{Resources occur in same
habitat}\label{resources-occur-in-same-habitat}

In this first scenario (equation \ref{eq:2}), the abundance of resources
at equilibrium are equivalent when both consumers and resources are
present (\(\hat R = \hat R_1 = \hat R_2\)), and are determined by the
following equation (derived in Appendix S2):

\begin{equation} \label{eq:5}
  \hat{R}=\frac{1}{a_{1,j}+a_{2,j}}\cdot\frac{m}{e}
\end{equation}

A key determinant of resource abundance in this scenario is the
consumer's total attack rate, \(a_{1,j}+a_{2,j}\). Therefore, the effect
of character displacement on food-web dynamics depends on how the shape
of the tradeoff function influences the evolution of consumer attack
rates.

I found that the shape of the tradeoff function qualitatively affects
the relationship between character displacement and resource abundances
in this scenario (fig. \ref{fig:plot_fig3} middle left). For example, if
consumer's are constrained by a linear tradeoff (blue lines), then there
is no net change in total attack rate (fig. \ref{fig:plot_fig3} top
left) and character displacement has no effect on resource abundances
(fig. \ref{fig:plot_fig3} middle left). If the tradeoff is concave down
(green lines), then resource abundances can actually increase under
character displacement (fig. \ref{fig:plot_fig3} middle left). This is
because the total attack rate of consumers is maximized at intermediate
values (\(a_{1,j}=a_{2,j}\)) and decreases as consumers diverge (fig.
\ref{fig:plot_fig3} top left). When the tradeoff is concave up (orange
lines), character displacement suppresses resource abundances due to the
increase in total attack rates (fig. \ref{fig:plot_fig3} top left).
Although the equation I derived for resource abundances was for the
scenario where both consumers and both resources were present, it
accurately predicts the abundance of resources when a single consumer
reaches its evolutionary stable strategy (ESS; triangles on respective
colored lines in fig. \ref{fig:plot_fig3} middle left). This is because
a single consumer evolves to be a generalist that has equal attack rates
on each resource (triangles at 0.5 along x-axis in fig.
\ref{fig:plot_fig3} left panels), resulting in equivalent resource
abundances.

The effect of character displacement on resources corresponds to its
impact on food-web stability. For example, when character displacement
decreases resource abundances (orange points in fig. \ref{fig:plot_fig3}
middle left), there is a decrease in food-web stability (fig.
\ref{fig:plot_fig3} bottom left). Character divergence may not affect or
even increase food-web stability (blue and green lines in fig.
\ref{fig:plot_fig3} bottom left); however, evolution does not favor
strong divergence in these scenarios (blue and green points in fig.
\ref{fig:plot_fig3}), which dampens these contingent effects. Note that
the dip in stability occurs when both consumers evolve to be
generalists, a situation that is not favored in any of the foraging
scenarios we examined (fig. \ref{fig:plot_fig3} bottom).

\subsection{Resources occur in different
habitats}\label{resources-occur-in-different-habitats}

In the second foraging scenario (equation \ref{eq:3}), I again see that
resource abundances are equivalent when both consumers and resources are
present (\(\hat R = \hat R_1 = \hat R_2\)), but are now determined by
the following equation (derived in Appendix S3):

\begin{equation} \label{eq:6}
  \hat{R}=\frac{1}{w_{1,j}a_{1,j}+w_{2,j}a_{2,j}}\cdot\frac{m}{e}
\end{equation}

This equation implies that if consumers evolve to become specialists on
resources that occur in their preferred habitat (e.g. \(w_{1,j}>0.5\)
and \(a_{1,j}>a_{2,j}\)), then the effective attack rate of consumers
(\(w_{1,j}a_{1,j}+w_{2,j}a_{2,j}\)) will always increase, regardless of
the tradeoff (fig. \ref{fig:plot_fig3} top right). Thus, character
divergence always results in resource suppression (fig.
\ref{fig:plot_fig3} middle right). Note that the shape of the tradeoff
can modify the effect of character displacement. This is not so much due
to the tradeoff affecting the magnitude of displacement (it does, but
the effect is minor), but because the form of the tradeoff affects
resource abundances when a single consumer has reached an ESS (triangles
in fig. \ref{fig:plot_fig3} middle right). In contrast, resource
abundances reach a similar value when consumers evolve in the presence
of a competitor (circles in fig. \ref{fig:plot_fig3} middle right),
because character displacement tends to reach a constraint of complete
specialization. It is worth noting that resource abundances are
consistently higher at the single consumer ESS compared to the
predictions I derived for when both consumers are present (deviation of
triangles from respective colored lines in fig. \ref{fig:plot_fig3}
middle right). This is because consumers actually evolve to be slightly
specialized on the resources that occur in their non-preferred habitat
(deviation of triangles from 0.5 along x-axis in fig.
\ref{fig:plot_fig3} right panels).

As seen previously, the effect of character displacement on resource
abundances qualitatively corresponds to its effect on food-web stability
(fig. \ref{fig:plot_fig3}). Specifically, character divergence decreases
food-web stability, regardless of the tradeoff in attack rates. This is
not simply a consequence of having an additional consumer in the system,
but emerges from the eco-evolutionary feedback between character
displacement and resource suppression (fig. \ref{fig:plot_fig3}). For
example, when the tradeoff is concave up (orange), the initial
two-consumer food web (small circle) is more stable than when there is
only one consumer (small triangle); however, this pattern switches by
the end of the eco-evolutionary simulation (large points).

\begin{figure}
\centering
\includegraphics{Figure_3.pdf}
\caption{\label{fig:plot_fig3}\textbf{Effect of character displacement
on food-web dynamics under different evolutionary tradeoffs and foraging
scenarios.} Lines show predicted values when both consumers and
resources are present. Different line colors correspond to different
tradeoffs in attack rates (green, \emph{n} = 1.15; blue, \emph{n} = 1;
orange, \emph{n} = 0.85). Large circles (two consumers) and triangles
(one consumer) correspond to the end points of the eco-evolutionary
simulation for \emph{C}\textsubscript{1}, whereas as small shapes
correspond to the starting points (only in stability panels). In both
foraging scenarios, feeding rates increase linearly with resource
abundance, but the equation for effective attack rate is different.}
\end{figure}

\subsection{Adding a more realistic functional
response}\label{adding-a-more-realistic-functional-response}

In the third foraging scenario (equation \ref{eq:4}), I observed the
same general effect of character divergence as the previous scenario
(resources in different habitats, but linear functional response). This
is because resource abundances at equilibrium are governed by a similar
dynamic (derived in Appendix S1):

\begin{equation} \label{eq:7}
  \hat{R}=\frac{1}{w_{1,j}a_{1,j}+w_{2,j}a_{2,j}}\cdot\frac{m}{e-hm}
\end{equation}

And since evolution favors consumer divergence onto their preferred
resources (see Appendix S1), the effective attack rate of consumers
(\(w_{1,j}a_{1,j}+w_{2,j}a_{2,j}\)) will always increase, resulting in
lower resource abundances and decreased food-web stability (Appendix
S4).

In the first two foraging scenarios, character displacement influences
food-web stability, but all of the food webs ultimately return to a
stable equilibrium (because \(-\lambda_{max}>0\), see Appendix S2-3). In
this more realistic model, however, whether the food web is locally
stable depends on consumer and resource parameters. Specifically, I
found that the two-consumer food web will transition from having a
locally stable equilibrium to a limit cycle under the following
conditions (derived using Routh-Hurwitz criteria in Appendix S1):

\begin{equation} \label{eq:8}
  w_{1,j}a_{1,j}+w_{2,j}a_{2,j} > \frac{e+hm}{hK(e-hm)}
\end{equation}

Thus, character displacement always pushes the food web toward an
unstable structure in this more realistic foraging scenario (fig.
\ref{fig:bifur_plot}). Note that I stopped the simulation in the
four-species food web once it became locally unstable. I do not simulate
beyond this point as this would require making assumptions about the
dynamics of mutant consumers in variable environments, which is beyond
the scope of this work.

\begin{figure}
\centering
\includegraphics{bifurcation_plot.pdf}
\caption{\label{fig:bifur_plot}\textbf{Character divergence creates an
unstable food web}. Lines illustrate the effects of character divergence
across the range of specialization, while the points are the results of
an eco-evolutionary simulation. Note that I increased the total
investment in attack rates (\emph{A} = 3.3) to create a scenario that
could result in an unstable food web.}
\end{figure}

\subsection{Robustness to consumer
asymmetry}\label{robustness-to-consumer-asymmetry}

The previous analytical results and simulations make a strong assumption
that competing consumers start off as perfect mirror images of each
other (i.e.~symmetry). Yet, theory indicates a predictable asymmetry
between initial consumer attack rates. This predictable asymmetry
emerges from a process of community assembly where a single consumer
invades a system, reaches an ESS as a generalist, followed by the
invasion of a second, more specialized, consumer. This theoretical
scenario has been hypothesized as the sequence of events leading to
character displacement in threespine stickleback in small coastal lakes
of British Columbia \citep{Schluter1992, Schluter2000}.

To test whether my results were robust to this asymmetry, I used the
evolved attack rates at the end of the simulations with one consumer as
the starting values for one of the two consumers. I did this for all
foraging scenarios and tradeoffs previously examined. I found that my
previous inferences are robust to including consumer asymmetry across
different foraging scenarios and tradeoffs (Appendix S4).

\section{Discussion}\label{discussion}

\subsection{Resource Abundances}\label{resource-abundances}

One of the criteria used to demonstrate character displacement is that
``sites of sympatry {[}two consumers{]} and allopatry {[}one consumer{]}
should not differ greatly in food, \ldots{}'' \citep{Schluter1992}. In
contrast, my results suggest that ecological character displacement
causes predictable differences in resource abundances. In fact, the
ecological and evolutionary scenarios that favored the greatest
character divergence always decreased the abundance of resources. For
example, the magnitude of character displacement was highest when mobile
consumers competed for resources that occur in different habitats. An
empirical example of this is threespine stickleback that have diverged
into limnetic and benthic species with specialized traits to forage on
zooplankton and benthic invertebrates, respectively
\citep{Schluter1992, Schluter2000}. These two resource types occur in
distinct zones of the lake (pelagic and littoral), therefore stickleback
must move between these zones when foraging for resources.
Interestingly, a disproportionate number of the documented cases of
character displacement involve carnivores \citep{Schluter2000} that are
larger, and likely more mobile, than their resources \citep{McCann2005},
suggesting many cases of ecological character displacement may result in
lower resource availability.

Similarly, the evolutionary tradeoff that favored character divergence
resulted in lower resources across all foraging scenarios. Although data
on the shape of the tradeoff in consumer foraging traits is scarce, two
classic examples of character displacement, Darwin's finches and
threespine stickleback, both appear to exhibit a tradeoff where extreme
trait values increase the net foraging rate of consumers
\citep{Schluter1985, Arnegard2014}. While it is theoretically possible
that character displacement does not alter (or even increase) resource
abundances, this was limited to the simplest, and arguably least
realistic, foraging scenario and under tradeoffs that did not favor
large displacements, and thus less likely to detect in nature. Taken
together, my results call for empirical work to test these clear
theoretical predictions and suggest a revision is needed for one of the
criteria used to demonstrate character displacement.

\subsection{Food-web Stability}\label{food-web-stability}

My most striking result was that character divergence made food webs
less resilient to perturbations. In fact, under the most realistic
foraging scenario, character divergence can even result in an unstable
food web. The mechanism underlying this destabilization is quite
general. Character divergence generally increases the strength of
consumer-resource interactions, but does not alter the strength of
intraspecific interactions. This relative increase in interspecific
interactions, combined with the natural oscillatory tendency of
consumer-resource dynamics \citep{Lotka1925, Volterra1926}, creates a
food-web structure that is less resilient to perturbations
\citep{Chesson2008, Rip2011, McCann2011}.

Interestingly, the ecological conditions that favor character divergence
are those that are already the least resilient to perturbations. For
example, \citet{McPeek2019} showed that character divergence is favored
in food webs that are either highly productive, easy to find and capture
resources, or under weak abiotic stress. This corresponds to higher
values of \(K\) (productivity) or \(A\) (investment in attack rates), or
lower values of \(m\) (abiotic stress). Each of these corresponding
changes decrease food-web resilience, as they increase the strength of
consumer-resource interactions relative to intraspecific interactions.
For example, increasing productivity reduces intraspecific competition
in resource populations while increasing the flux of energy to
consumers, resulting in the paradox of enrichment
\citep{Rosenzweig1971}. Similarly, higher feeding rates or lower
consumer mortality both increase the relative strength of
consumer-resource interactions, which predictably destabilizes food webs
\citep{Rip2011, McCann2011}. This suggests that the most dramatic
examples of character divergence will not only occur in, but also cause,
the least stable food-web structures.

A handful of empirical patterns support the hypothesis that character
divergence decreases food-web resilience. For example, a single species
of threespine stickleback lives in hundreds of small coastal lakes of
British Columbia, but the species pair, where character divergence has
resulted in specialized limnetic and benthic species, are only known
from six lakes in four independent watersheds
\citep{Schluter1992, Schluter2000}. Perhaps many lakes had a species
pair in the past, but have lost a species due to a less resilient
food-web structure \citep{Borrelli2015a, Borrelli2015b}. The species
pair are known to be vulnerable to perturbations, as they have gone
extinct in two of the six lakes after the introduction of nonnative
species \citep{Hatfield2001, Taylor2006, Rudman2016}. The vulnerability
of the stickleback system also corresponds with the fact that aquatic
food webs have several properties that make them less resilient to
perturbations, such as higher productivity and more efficient energy
transfer to consumers \citep{Rip2011}. Detecting the ghost of
competition past \citep{Connell1980} may be quite difficult, but it
could be possible with recent advances in genomics. For example,
\citet{Feulner2018} detected genomic signatures of hybridization in
sympatric whitefish species following periods of eutrophication. Perhaps
solitary stickleback in some lakes retain genomic signatures of having
been a habitat specialist in the past.

My results contrast, but do not necessarily contradict, the notion from
coexistence theory that character displacement contributes to species
coexistence \citep{Lawlor1976}. Rather than studying resilience,
coexistence theory usually studies the mutual ability of consumers with
different phenotypes to invade when rare \citep[mutual
invasibility,][]{Chesson2000}. In the context of character displacement,
a shortcoming of this mutual invasibility measure is that it does not
allow a comparison between food webs with and without a competing
consumer. Such comparisons are necessary for inferring the effects of
character displacement, a point that has been made clear in the criteria
to demonstrate character displacement
\citep{Schluter1992, Schluter2000}. Although the addition of a consumer
to a food web can decrease its resilience in the absence of evolution
\citep{May1973}, my results are primarily driven by an eco-evolutionary
feedback between consumer evolution and resource abundances.

\subsection{Caveats}\label{caveats}

Although I model the indirect effects of coevolution between consumers,
I do not account for potential coevolution between consumers and
resources. In the context of my model, I would expect prey to evolve
traits that reduce consumer attack rates. Thus, prey evolution would act
to counter the effects of character displacement on resource abundance
and food-web stability. Note that this does not negate my general
conclusion that ecological character displacement decreases resource
abundances and stability; however, this process may itself create
another eco-evolutionary feedback between consumers and resources. This
may actually help maintain dramatic examples of character divergence and
prevent them from destabilizing systems, because it allows consumer
traits to become decoupled from their attack rate. Examining this
decoupling would be ideal in a quantitative genetic model that can
explicitly track trait dynamics, but it would not fundamentally change
the conclusions presented here.

Another potential caveat is that I explored my model in a setting that
makes many assumptions about resource and consumer symmetry (but see
consumer asymmetry section). Prior work has shown that allowing for
resource asymmetry, for example, may decrease the magnitude of character
displacement \citep{Abrams1986}. This is because this asymmetry creates
an asymmetry in resource abundances, which dampens the effects of
character divergence. Again though, this may dampen the effects of
divergence, but should not change the qualitative relationship we
observed of ecological character displacement decreasing resource
abundances and food-web stability.

>>>>

\usepackage{fullpage}
\linespread{1.5}
\usepackage{lineno}
\usepackage{caption}
\captionsetup[figure]{font=small,labelformat=empty}

@@

\begin{center} \LARGE Appendix S4: \\ Ecological character displacement destabilizes food webs \end{center}

@@

\emph{Elements}: Figures S1, S2, and S3.

@@

\begin{figure}
\centering
\includegraphics{Figure_ECD_McCann.pdf}
\caption{Figure S1: \textbf{Effect of character displacement when
consumers exhibit a more realistic functional response.} Different line
colors correspond to different tradeoff values (green, \emph{n} = 1.15;
blue, \emph{n} = 1; orange, \emph{n} = 0.85). Large circles (two
consumers) and triangles (one consumer) correspond to the end points of
the eco-evolutionary simulation for \emph{C}\textsubscript{1}, whereas
as small shapes correspond to the starting points (only in stability
panel). This figure illustrates how, regardless of the tradeoff,
character displacement increases the effective attack rate of consumers.
This always resulted in a suppression of resource abundances and
concomitant decrease in food-web stability. Initial parameter and state
variables are given in the main text.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{asymm_figs.pdf}
\caption{Figure S2: \textbf{Robustness to consumer asymmetry}. Lines
show predicted values when both consumers and resources are present.
Different line colors correspond to different tradeoffs in attack rates
(green, \emph{n} = 1.15; blue, \emph{n} = 1; orange, \emph{n} = 0.85).
Large circles (two consumers) and triangles (one consumer) correspond to
the end points of the eco-evolutionary simulation for
\emph{C}\textsubscript{1}, whereas as small shapes correspond to the
starting points (only in stability panels). In both foraging scenarios,
feeding rates increase linearly with resource abundance, but the
equation for effective attack rate is different. The similarity with
fig. 3 (main text) shows that adding an asymmetry in consumer attack
rates does not alter the conclusions reported in the main text. If
anything, it is more likely that an asymmetry will push the system
toward the boundary of stability (note position of large circles in
lower left stability panel). Details of the consumer asymmetry
simulation are given in the main text.}
\end{figure}

\begin{figure}
\centering
\includegraphics{Figure_ECD_McCann_asymm.pdf}
\caption{Figure S3: \textbf{Robustness to consumer asymmetry when
consumers exhibit a more realistic functional response.} Different line
colors correspond to different tradeoff values (green, \emph{n} = 1.15;
blue, \emph{n} = 1; orange, \emph{n} = 0.85). Large circles (two
consumers) and triangles (one consumer) correspond to the end points of
the eco-evolutionary simulation for \emph{C}\textsubscript{1}, whereas
as small shapes correspond to the starting points. Panel (A) shows how,
regardless of the tradeoff, character displacement increases the
effective attack rate of consumers. This always resulted in a
suppression of resource abundances (B) and concomitant decrease in
food-web stability (C). The similarity with fig. S1 shows that adding an
asymmetry in consumer attack rates does not alter the conclusions
reported in the main text. Details of the consumer asymmetry simulation
are given in the main text.}
\end{figure}
